<!--
%\VignetteEngine{knitr::docco_linear}
%\VignetteIndexEntry{An Introduction to the pegr package}
-->

An Introduction to the **Pegr** Package
=======================================


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center")
```


Introduction
------------


The  **pegr** package is a package containing a parsing expressing grammer generator, 
together with some tools for managing and debuging sets of rules. Pegr is a recursive decent parser, 
using memoizaton to improve efficiency. The Pegr grammar is the creation of ***Brian Ford***. 
This document is divided up into two major sections, the first just describes the grammar, the second gives a walkthrough on usage within **pegr** 

### The Grammar
The grammar consists of a set of rules ( a rule is also called a definition)
* **Rules** A rule consists of a ruleid, an arrow, and a parsing expresssion.  The ruleid is also called a rule name or a non-termial.
A rule a has the form
$$ RuleId \leftarrow ParsingExpression$$
* **Parsing Expressions**  are built from atoms (aka. literals or terminals) rules, and operators. In pegr, each parsing expression returns a result consisting of three items:
1. A *status* (TRUE for a match, i.e. success, FALSE for no match, i.e. failure )
2. A *position* recording what of the original text was consumed.
3. A *value*, actually a list of one one more values associated with that expression.

#### The following describes the components of a parsing expression:

* **Atoms** are terminal expressions. Atoms consist of literals, ranges, or the dot wild card. Atom may be a single letter 'a' , a word "dog"" or a ranges such as [a-z]. When an atom is matches, the return status is TRUE and a singleton list of value of the match is returned
* **Sequences** A collections of expressions seperated by a blank spaces is called a *sequence*. A sequence is of the form:
   $$e_1 \ e_2\  e_3 \ ... \ e_n$$ where each $e_n$ is an expression. 
For a sequence to match, each component must match successively. For example $$.\ 'a'$$ recognizes any symbol followed by the lower case a. A sequence may be thought of as an and operator. The value returned by sequence is a list consisting of all the values of the individuals components which built up the sequence. 
* **Ordered choice operators**  *Ordered choices operators* are  of the form 
   $$e_1 / e_2 / ... / e_n$$
   Ordered choice returns the value of the first match (if any). If there is no match, the status becomes false, and the expression is rejected. For example, 
   $$'a' \ \ 'b' \ / \ 'c'$$
will match "ab" and "ac".
* **Zero or More operator** The zero or more operator will match if zero or more of occurances of the preceding expression are present. The zero or more operator representation is of the form 
  $$e*$$
  For example $$'a'*$$ will match "aaa" and "".
* **One or More operator** The *one or more operator* will match if zero or more of occurances of the preceding expression are present. The *one or more operator* representation is of the form 
  $$e+$$
  For example $$'a'+$$ will match "aaa" but not "".
* **Optional operator** The *optional operator*, or at *most one operator* will match if zero or one of occurances of the preceding expression are present. The optional operator representation is of the form 
  $$e?$$
  For example $$'a'?$$ will match "a" and "", but not "aaa".

* **And Predicate** The *And Predicate* is a look-ahead operator which will match if the next element is a match. It is look ahead in the sense that it does not consume the next expression, it only rejects the current expression if the next expression does not follow. The *And Predicate* is of the form 
  $$e\&$$
  For example $$'a'\ \& \ 'b'$$ will match "ab" but will return only list(atom="a")

* **Not predicate** The *Not Predicate* is also a look-ahead operator but unlike the *And Predicate*, the *Not Predicate* will fail to match if the next element is a match.  The *Not Predicate* is of the form 
  $$!e$$
  For example $$'a'\ ! \ 'b'$$ will match "ax" but will return only list(atom="a")


Usage Specifics
----------------------------
### Parser
A pegr parser is created by a call to new.parser()
```{r methods}
library(pegr)
parser <- new.parser()
```

### Rules
* Rules may be added to a pegr parser by a call to **add_rule(parser, rule)**
```{r add_rule}
add_rule(parser, "Any<-.")
add_rule(parser, "A<-'a'")
add_rule(parser, "B<-'b'")
add_rule(parser, "C<- A (B / C) ")
add_rule(parser, "D<- D") #bad rule: will produce infinite recursion   
```

* The ids or names of all rules in a pegr parser may be retrieved by **rule_ids**
```{r rule_ids}
rule_ids(parser)
```

* Rules contained in a pegr parser may be annotated (commented) by a call to **set_description("ruleid", annotation)**
```{r set_description}
set_description(parser, "Any", "Accepts any character")
set_description(parser, "A", "Accepts a")
set_description(parser, "B", "Accepts b")
set_description(parser, "C", "Accepts string of a's terminated by a b")
set_description(parser, "D", "A very bad rule")
```

* Rules may be removed from a pegr parser by a call to **delete_rule(parse, ruleid)**
```{r delete_rule}
delete_rule(parser, "D")
rule_ids(parser)
```

### Actions
Actions may be attached to an existing rule contained pegr parser by a call to **add_action(parse, action)**
An action may be one of two forms:
* **An r function**:
```{r add_action_function}
rule_ids(parser)
```
* **A character string**. This character string should contain code to be executed as the body of a function, where the
input is a list v, and the return value is also a list.
```{r add_action_char_string}
set_action(parser, "A", "list('A')") #turn a lower case a to an upper case A
set_action(parser, "C", "list(paste(v,collapse=''))") #paste all the characters together
```

### Parsing
Parsing is the act of applying a rule to a string to be parsed and returning a result. 
The rule is specified its ruleid, a character string giving the rules name. That rule becomes the
root of the ensuing parse.
* Parsing is accomplished by
** apply_rule(parser, ruleid, exe=FALSE,, debugTree=FALSE)**. 
```{r apply_rule}
apply_rule(parser, "A", "a" )->res.A  
apply_rule(parser, "C", "aab")->res.AAB
apply_rule(parser, "C", "baab")->res.BAA
```
* Status of a parsing attempt is either **TRUE** for success or **FALSE** for failure. 
The status may be obtained by **status(res)**
```{r status.ResA}
c(status(res.A) ,status(res.AAB), status(res.BAA) )
```
* The characters consumed by parsing attempt can be gotton from **consumed(res)**
```{r consume}
c(consumed(res.A),  consumed(res.AAB), consumed(res.BAA))
```

* Return values of a parsing are lists, which can be gotton from **value(res)**
By default, actions defined by **set_action** are not executed.
```{r valueF}
# here exe is false, so no action is taken
apply_rule(parser, "C", "aab")->res.AAB 
# so the return is a list of the returns of the component atoms
value(res.AAB)
```
To execute the actions, supply **exe=TRUE** as a parameter:
* Return values of a parsing are lists, which can be gotton from **value(res)**
```{r valueT}
#here exe is true, so action is taken
apply_rule(parser, "C", "aab", exe=TRUE)->res.AAB 
# action A capitalizes, action C pastes together
value(res.AAB) 
```




### Debugging
Debugging the logic of a grammar can be accomplished testing the component rules starting with the leaves and building up  Also, it's probably better, to start with actions disabled  to see if the general flow is correct first. To add in the anaylsis of a given parse there are two 
visualizations tools:
* **tree**: 
* **plot**: 
Thesse  visualizaton tools are discussed in the next section

### Visualization
Visualization consists of showing the nodes visited (and their values) during a parse of an input. This requires that the
**debugTree** flag be set to **TRUE** in the invocation of **apply_rule**. 
```{r debugRuleT}
# Here debugTree is set to True
apply_rule(parser, "C", "aab", exe=TRUE, debugTree=TRUE)->res.AAB 
# also, since exe isTrue, so action is taken
# action A capitalizes, action C pastes together
value(res.AAB) 
```

#### Ways to Visualize
There are two methods of visualization:
* **Tree**: Prints the tree to console
```{r debugRuleTree}
tree(res.AAB) 
```

* **Plot**: Plots the tree to the graphics device
```{r debugRulePlot}
#plot only the ruleids (names)
plot(res.AAB) 
#plot only the ruleids (names and inputs)
plot(res.AAB, show=c("names", "args")) 
#plot only the values
plot(res.AAB, show="vals")
#plot all
plot(res.AAB, show="all")
```



